Intersection Types, Sub-formula Property, and the Functional Character of the Lambda Calculus

### 2018-08-09 DenotSem/determ.tex
check
longer proof for var case

We shall represent variables with de Bruijn indices, so an environment
$\rho$ is a sequence of types.  The denotational semantics of the CBV
$\lambda$-calculus is defined as follows.
\begin{align*}
  \ESEM{n}\rho &= \UP{n} \\
  \ESEM{x}\rho &=  \IF x < |\rho| \THEN \UP{\rho_k} \ELSE \emptyset \\
  \ESEM{\lambda e }\rho &= \{ A \mid \WF{A} \text{ and } \FSEM{A}(e,\rho) \} \\
  \ESEM{e_1\app e_2}\rho &= \left\{ C\, \middle| \exists A,B.\;
      \begin{array}{l}
      A \in \ESEM{e_1}\rho,
      B \in \ESEM{e_2}\rho,\\
      A <: B \to C, \text{ and } \WF{C}
      \end{array}
        \right\} \\
  \ESEM{f(e_1,e_2)}\rho &=
      \left\{ C\, \middle| \begin{array}{l}
       \exists A,B,n_1,n_2.\; A \in \ESEM{e_1}\rho, B \in \ESEM{e_2}\rho,\\
      \qquad A <: n_1, B <: n_2, \SEM{f}(n_1,n_2) <: C, \WF{C} 
      \end{array} \right\} \\
  \ESEM{\IF e_1 \THEN e_2 \ELSE e_3}\rho &=
    \left\{ B\, \middle| \exists A.\;
    \begin{array}{l}A \in \ESEM{e_1}\rho, A <: n,\\
           n = 0 \Rightarrow B \in \ESEM{e_3}\rho,\\
           n \neq 0 \Rightarrow B \in \ESEM{e_2}\rho
    \end{array}
    \right\}
    \\[2ex]
    \intertext{where}
   \FSEM{n}e\rho &= \mathit{false} \\
   \FSEM{A \sqcap B}e \rho &= \FSEM{A}e\rho \text{ and } \FSEM{B}e\rho\\
   \FSEM{A \to B}e \rho &= B \in \ESEM{e} (A, \rho)
\end{align*}

@@

Now we can move on to lemmas about the denotational semantics that
will be needed to prove the Join Theorem. \\

\noindent \textbf{Lemma} (Weakening)
\begin{enumerate}
\item If $\FSEM{A}e \rho_1$, $\rho_1 <: \rho_2$ and
  $(\forall B \rho_1 \rho_2,\; B \in \ESEM{e}\rho_1, \rho_2 <: \rho_1
  \Rightarrow B \in \ESEM{e}\rho_2)$, then
  $\FSEM{A}e \rho_2$.
\item If $A \in \ESEM{e}\rho_1$ and $\rho_2 <: \rho_1$,
  then $A \in \ESEM{e}\rho_2$.
\end{enumerate}
(Part 1 is proved by induction on $A$. Part 2 is proved by induction
on $e$ and uses part 1.)

\noindent \textbf{Lemma} (Consistent Application) \\
%
If $A_1 \sim A_2$, $B_1 \sim B_2$, $A_1 <: B_1 \to C_1$,
$A_2 <: B_2 \to C_2$, and all these types are well formed,
then $C_1 \sim C_2$.

>>>>

We shall represent variables with de Bruijn indices, so an environment
$\Gamma$ is a sequence of types.  The denotational semantics of the CBV
$\lambda$-calculus is defined as follows.
\begin{align*}
  \ESEM{n}\Gamma &= \UP{n} \\
  \ESEM{x}\Gamma &=  \IF x < |\Gamma| \THEN \UP{\Gamma[k]} \ELSE \emptyset \\
  \ESEM{\lambda e }\Gamma &= \{ A \mid \WF{A} \text{ and } \FSEM{A}(e,\Gamma) \} \\
  \ESEM{e_1\app e_2}\Gamma &= \left\{ C\, \middle| \exists A,B.\;
      \begin{array}{l}
      A \in \ESEM{e_1}\Gamma,
      B \in \ESEM{e_2}\Gamma,\\
      A <: B \to C, \text{ and } \WF{C}
      \end{array}
        \right\} \\
  \ESEM{f(e_1,e_2)}\Gamma &=
      \left\{ C\, \middle| \begin{array}{l}
       \exists A,B,n_1,n_2.\; A \in \ESEM{e_1}\Gamma, B \in \ESEM{e_2}\Gamma,\\
      \qquad A <: n_1, B <: n_2, \SEM{f}(n_1,n_2) <: C, \WF{C} 
      \end{array} \right\} \\
  \ESEM{\IF e_1 \THEN e_2 \ELSE e_3}\Gamma &=
    \left\{ B\, \middle| \exists A.\;
    \begin{array}{l}A \in \ESEM{e_1}\Gamma, A <: n,\\
           n = 0 \Rightarrow B \in \ESEM{e_3}\Gamma,\\
           n \neq 0 \Rightarrow B \in \ESEM{e_2}\Gamma
    \end{array}
    \right\}
    \\[2ex]
    \intertext{where}
   \FSEM{n}e\Gamma &= \mathit{false} \\
   \FSEM{A \sqcap B}e \Gamma &= \FSEM{A}e\Gamma \text{ and } \FSEM{B}e\Gamma\\
   \FSEM{A \to B}e \Gamma &= B \in \ESEM{e} (A, \Gamma)
\end{align*}

@@

The next two lemmas follow from the Consistency and Subtyping Lemma
and help prepare to prove the case for application in the Join
Theorem.

\noindent \textbf{Lemma} (Application Consistency) \\
%
If $A_1 \sim A_2$, $B_1 \sim B_2$, $A_1 <: B_1 \to C_1$,
$A_2 <: B_2 \to C_2$, and all these types are well formed,
then $C_1 \sim C_2$.\\
(This lemma is proved directly, without induction.)\\

\noindent \textbf{Lemma} (Application Intersection) \\
%
If $A_1 <: B_1 \to C_1$, $A_2 <: B_2 \to C_2$, $A_1 \sim A_2$,
$B_1 \sim B_2$, and $C_1 \sim C_2$, then
$(A_1\sqcap A_2) <: (B_1 \sqcap B_2) \to (C_1 \sqcap C_2)$.\\
(This lemma is proved directly, without induction.) \\

We move on to the Weakening Lemma, which will be needed to prove the
Join Theorem. \\

\noindent \textbf{Lemma} (Weakening)
\begin{enumerate}
\item If $\FSEM{A}e \Gamma_1$, $\Gamma_1 <: \Gamma_2$ and
  $(\forall B \Gamma_1 \Gamma_2,\; B \in \ESEM{e}\Gamma_1, \Gamma_2 <: \Gamma_1
  \Rightarrow B \in \ESEM{e}\Gamma_2)$, then
  $\FSEM{A}e \Gamma_2$.
\item If $A \in \ESEM{e}\Gamma_1$ and $\Gamma_2 <: \Gamma_1$,
  then $A \in \ESEM{e}\Gamma_2$.
\end{enumerate}
(Part 1 is proved by induction on $A$. Part 2 is proved by induction
on $e$ and uses part 1.) \\

Now for the main event, the proof of the Join Theorem!
Though we'll rebrand it as the Meet Theorem.\\

\noindent \textbf{Theorem} (Meet)\\
If $A_1 \in \ESEM{e}\Gamma_1$, $A_2 \in \ESEM{e}\Gamma_2$,
both $\Gamma_1$ and $\Gamma_2$ are well formed,
and $\Gamma_1 \sim \Gamma_2$, \\
then $A_1 \sqcap A_2 \in \ESEM{e}(\Gamma_1\sqcap\Gamma_2)$. \\
%
\textbf{Proof}
We proceed by induction on $e$.
\begin{itemize}
\item Case $e=x$:
  We have $\Gamma_1[k] <: A_1$ and $\Gamma_2[k] <: A_2$.

  nts. $\Gamma_1[k] \sqcap \Gamma_2[k] <: A_1 \sqcap A_2$
\end{itemize}
